---
  output: bookdown::html_document2
  html_document: default
  word_document: default
---
Activity monitoring data
=========================

```{r setoptions,include=FALSE}
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,results="hide",
                      fig.keep="all")
```
## Introduction

It is now possible to collect a large amount of data about personal movement 
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone 
Up. These type of devices are part of the “quantified self” movement – 
a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are
tech geeks. But these data remain under-utilized both because the raw data are 
hard to obtain and there is a lack of statistical methods and software for 
processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. 
The data consists of two months of data from an anonymous individual collected 
during the months of October and November, 2012 and include the number of steps 
taken in 5 minute intervals each day.

```{r readprocdata}
fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp<-tempfile() # create a temporary file
download.file(fileUrl,temp) #download file into temporary location
activity<-read.csv(unz(temp,"activity.csv"))
library(dplyr)
activityGrp<-group_by(activity,date)
activityDaily<-summarise(activityGrp,dailystep=sum(steps,na.rm=TRUE))
source("formatNum.R")
```

### Total number of steps taken per day
The median number of steps taken each day was 
**`r formatNum(median(activityDaily$dailystep),0)`**, whilst the mean was
**`r formatNum(mean(activityDaily$dailystep),0)`**, when missing values were not
considered. A histogram displays the distribution of total steps in a day below 
.Note that a considerable number of observations had missing 
values, denoted in the 0 bar.

```{r plothist,fig.cap="Histogram of total number of steps taken per day"}
library(ggplot2)
qplot(dailystep,data=activityDaily,ylab="Frequency",xlab="Total steps in a day",
      binwidth=2500,geom="histogram",fill=I("springgreen2"),col=I("springgreen3"))
```

### Daily activity pattern
```{r groupByInt}
activityGrp<-group_by(activity,interval)
activityInt<-summarise(activityGrp,avgstep=mean(steps,na.rm=TRUE))
```
The maximum number of steps on average was observed on the **`r activityInt$interval[which.max(activityInt$avgstep)]`th** interval.
```{r plotscatt,fig.cap="Time series plot of average number of steps versus interval"}
qplot(interval,avgstep,data=activityInt,ylab="Average number of steps during 
      interval",xlab="Interval",col=I("steelblue1"),geom="line")
```
### Missing values
``` {r missingcalc}
totalrows<-nrow(activity)
completerows<-sum(complete.cases(activity))
missingrows<-totalrows-completerows
library(Hmisc)
imputedsteps<-impute(activity$steps,fun=mean(activity$steps,na.rm=TRUE))
activityNew<-activity
activityNew$steps<-imputedsteps
activityGrp<-group_by(activityNew,date)
activityNewDaily<-summarise(activityGrp,dailystep=sum(steps,na.rm=TRUE))
```
The dataset contained many missing values. Out of 
**`r formatNum(totalrows,0)`** observations, only **`r formatNum(completerows,0)`**
were complete with the remaining **`r formatNum(missingrows,0)`** missing data 
on the number of steps.When missing values were imputed, the median number of 
steps taken each day was **`r formatNum(median(activityNewDaily$dailystep),0)`**,
whilst the mean was **`r formatNum(mean(activityNewDaily$dailystep),0)`**, both
values higher than estimated when excluding missing data.
```{r plothist2,fig.cap="Histogram of total number of steps taken per day with imputed data"}
qplot(dailystep,data=activityNewDaily,ylab="Frequency",xlab="Total steps in a day (imputed)", binwidth=2500,geom="histogram",fill=I("violetred"),col=I("violetred1"))
```
### Differences between weekdays and weekends
The plots below display the average number of steps at different intervals
during weekdays and weekends.
``` {r plotdaytype,fig.cap="Time series plot of interval versus average number of steps by type of day"}
library(lubridate)
activity$date<-ymd(activity$date)
activity$day<-weekdays(activity$date)
activity$dayType<-"Weekday"
activity$dayType[activity$day=="Saturday"]<-"Weekend"
activity$dayType[activity$day=="Sunday"]<-"Weekend"
activityByDayType<-aggregate(activity,by=list(activity$interval,activity$dayType),
                             FUN=mean,na.rm=TRUE)
library(tidyverse)
activityByDayType<-rename(activityByDayType,DayType=Group.2,Interval=Group.1)

qplot(Interval,steps,data=activityByDayType,facets=.~DayType,
ylab="Average number of steps during interval",xlab="Interval",
col=DayType,geom="line")
```